Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-09 paper/examples.tex
added mixed map and cut cut cut

Our first example, \texttt{dot}, appears in \cite{knoblock96}. \texttt{dot} is a
first-order, non-recursive function---precisely the sort of code studied in
prior work on pass separation for imperative languages. \texttt{dot} takes the
dot product of two three-dimensional integer vectors, where the first two
coordinates are first-stage, and the last coordinate is second-stage. 
%
\begin{lstlisting} 
1`type vec = ^int * ^int * $`2`int`

1`dot : vec * vec -> $int
fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
\end{lstlisting}
%
The body of \texttt{dot} is an \rmint\ term at world \bbtwo\ containing an
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

Our algorithm splits \texttt{dot} into the two functions:
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

@@

Thus, \texttt{qs1} performs $\Theta(n \log n)$ work per call, whereas
\texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
speedup over standard quickselect if we perform $\omega(\log n)$ queries on
the same list.

Note that the recursive \texttt{part} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.
%

\subsection {Mixed Map Combinator}

[working on this section!]

In \ref{sec:exampleQS}, we showed the output of splitting quickselect
applied to one list and $m$ queries.  Here, we generalize to mapping
quickselect onto a list of queries. Note that we have to define a
datatype for integer lists at the second stage, which is given by
\texttt{list2}.
\begin{lstlisting} 
1`atsignnext{`2`
  datatype list2 = Empty2 | Cons2 of int * list2
`1`}

map : (^list * $`2`int`1` -> $`2`int`1`)
      -> ^list * $`2`list2`1` -> $`2`list2`1`
fun map f (l, q) = next {`2`
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) =
                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
  in m prev{`1`q`2`}`1`}

val mapqss = map qss`
\end{lstlisting}
The \texttt{map} function splits into the following two functions:
\begin{lstlisting} 
2`datatype list2 = Empty2 | Cons2 of int * list2

`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
`2`fun map2 (f,()) ((l,q), p : tree) =
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
  in m q

`1`val (mapqss1,()) = map1 qs1
`2`val mapqss2 = map2 (qs2, ())`
\end{lstlisting}
%
Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
in the second pass, once per query \texttt{k}.

@@

Composable staged programs are particularly important the domain of real-time graphics.
This need arises because modern graphics architectures actually
\emph{require} that graphics computations be structured as a pipeline
of stages which perform increasingly fine-grained computations (e.g.,
per-object, per-screen region, per-pixel). Many computations in later
stages use the results from a single computation in an earlier
stage\,\cite{OpenGL4Spec}.

The standard way to program these graphics pipelines is to define one program (usually called a \emph{shader}) per stage.
In other words, the programmer is expected to manually split their multi-stage programs.
Yet manually partitioning programs by stage results in
complex code where invariants must hold across different stages and
local changes to the logic of one stage may require changes to that of
upstream stages to communicate required inputs through the pipeline. 
This harms composition and modularity.  
Graphics researchers therefore have suggested using mixed-staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic 
for the entire pipeline as a single staged function\,\cite{Foley:2011}.

As a language, \lang\ is well suited for specification of such functions,
and we give a simple example below.
In it, we consider a graphics pipeline program to be a
staged function that takes an object definition in the first stage (\texttt{object})
and a pixel coordinate in the second stage ($\fut$\texttt{coord}), 
and emits the color of the object at the specified pixel ($\fut$\texttt{color}). 
Given two such multistage functions, we then define a combinator
that multiplies their results pointwise in the second 
stage.\footnote{Although this is a general pointwise multiplication combinator, 
the variable names suggest a possible interpretations of the inputs and output:
the first input function calculates the albedo of an object (a measure of how much light it reflects),
the second input calculates the object's incoming light,
and so the output (the product of these terms) is a function that calculates 
the outgoing light from an object.}
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{  type coord = int * int
        type color = ...        }`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
  fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
\end{lstlisting}
%
A key strength of \lang\ is that it allows users to define combinators like this one as higher order functions.
Prior work\,\cite{Foley:2011} does not support higher order functions.

>>>>

Our first example, \texttt{dot}, appears in \cite{knoblock96}. \texttt{dot} is a
first-order, non-recursive function---precisely the sort of code studied in
prior work on pass separation for imperative languages. \texttt{dot} takes the
dot product of two three-dimensional vectors, where the first two
coordinates are first-stage, and the last coordinate is second-stage. 
%
\begin{lstlisting} 
1`type vec = ^int * ^int * $`2`int`
1`// dot : vec * vec -> $int
fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
\end{lstlisting}
%
The body of \texttt{dot} is an \rmint\ term at world \bbtwo\ containing an
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.
Our algorithm splits \texttt{dot} into the two functions:
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, though
they write the precomputation into a mutable cache.

@@

Note that the recursive \texttt{part} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.

As discussed in \ref{sec:overview}, \texttt{qs1} performs $\Theta(n \log n)$ expected work per call, 
whereas \texttt{qs2} performs $\Theta(\log n)$ expected work.  
This results in a net speedup over standard quickselect if we 
perform many (specifically, $\omega(\log n)$) queries on
the same list---precisely the topic of our next example.

\subsection {Mixed-Stage Map Combinator}

As a lambda calculus, one of the strengths of \lang\ is that it
can express combinators as higher order functions.
In this example, we consider just such a combinator: \texttt{tmap},
which turns a function of type
$\curr \mathtt{list} * \fut \mathtt{int} \to \fut \mathtt{int}$
into one of type
$\curr \mathtt{list} * \fut \mathtt{list2} \to \fut \mathtt{list2}$,
by mapping over the second argument.\footnote{
\lang\ doesn't have a way to ``share'' datatype declarations between stages, 
so we define \texttt{list2} to be a list of integers at the second stage.}
\begin{lstlisting} 
1`atsignnext{`2` datatype list2 = Empty2 | Cons2 of int * list2 `1`}

fun tmap (f : ^list * $`2`int`1` -> $`2`int`1`) (l : ^list, q : $`2`listliteraltwo`1`) = 
  next{`2` let fun m Empty2       = Empty2
              | m (Cons2(h,t)) = Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
        in m prev{`1`q`2`}`1`}
val mapqss = tmap qss`
\end{lstlisting}
Importantly, \texttt{tmap f} performs the first-stage part of \texttt{f} once
and second-stage part of \texttt{f} many times.
This was discussed in the context of partial evaluation in \ref{sec:topLevel}, 
but it is especially clear when we look at the output of splitting:
\begin{lstlisting} 
1`fun tmap1 f = (fn (l,()) => ((),#2 (f (l,()))),())
val (mapqss1,()) = tmap1 qs1

`2`datatype list2 = Empty2 | Cons2 of int * list2
fun tmap2 (f,()) ((l,q), p : tree) =
  let fun m Empty2 = Empty2
        | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
  in m q
val mapqss2 = tmap2 (qs2, ())`
\end{lstlisting}
%
Indeed, observe that the argument \texttt{f} of \texttt{qs1} is called only once,
whereas the corresponding argument \texttt{f} in \texttt{qs2} is evaluated once per element in \texttt{q}.

@@

Composable staged programs are particularly important the domain of real-time graphics.
This need arises because modern graphics architectures actually
\emph{require} that graphics computations be structured as a pipeline
of stages which perform increasingly fine-grained computations (e.g.,
per-object, per-screen region, per-pixel), where computations in later
stages use the results of an earlier
stage multiple times\,\cite{OpenGL4Spec}.

The standard way to program these graphics pipelines is 
to define one program (usually called a \emph{shader}) per stage.
In other words, the programmer is expected to write their 
multi-stage programs in an already split form.
This requirement results in complex code where invariants must hold across different stages and
local changes to the logic of one stage may require changes to that of upstream stages. 
This harms composition and modularity.  
Graphics researchers therefore have suggested using mechanisms like our stages
\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic,
including representing entire pipeline as a single multi-stage function\,\cite{Foley:2011}.

As a language, \lang\ is well suited for specification of such functions,
and we give a simple example below.
In it, we consider a graphics pipeline program to be a
staged function that takes an object definition in the first stage (\texttt{object})
and a pixel coordinate in the second stage ($\fut$\texttt{coord}), 
and emits the color of the object at the specified pixel ($\fut$\texttt{color}). 
Given two such multistage functions, we then define a combinator
that multiplies their results pointwise in the second 
stage.\footnote{Although this is a general pointwise multiplication combinator, 
the variable names suggest a possible interpretation of the inputs and output:
the first input function calculates the albedo of an object (a measure of how much light it reflects),
the second input calculates the object's incoming light,
and so the output (the product of these terms) is a function that calculates 
the outgoing light from an object.}
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{  type coord = int * int
        type color = ...        }`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
  fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
\end{lstlisting}
%
Prior work\,\cite{Foley:2011} lacks the ability to define such combinators,
because it does not support higher order functions.

### 2016-01-09 paper/implementation.tex
added mixed map and cut cut cut

\subsection {Proof Implementation}

We have encoded the type system, dynamic semantics, splitting algorithm, and output semantics in Twelf,
as well as all of the theorems of \ref{sec:semantics,sec:splitting}, and their proofs. 
Details can be found in the supplementary materials.

\subsection {Example Implementation}

\begin{abstrsyn}
We have a Standard ML implementation of the \lang\ language, a staged
interpreter, the splitting algorithm, and an interpreter for split programs.
This implementation extends the language described in \ref{sec:semantics} with 
\rmint{}s,
\rmbool{}s,
\texttt{let} statements,
$n$-ary sums and products,
datatype and function declarations, and
deep pattern matching (including $\next{}$ and $\pure{}$ patterns).
The code snippets in this paper are written in our concrete syntax, using 
these additional features when convenient.

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngr{`1`
  datatype list = Empty | Cons of int * list
  fun part (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
val 3`gr{`part3`}` = 3`gr{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

>>>>

\begin{abstrsyn}
We have encoded the type system, dynamic semantics, splitting algorithm, and output semantics in Twelf,
as well as all of the theorems of \ref{sec:semantics,sec:splitting}, and their proofs. 
We also have a Standard ML implementation of the \lang\ language, a staged
interpreter, the splitting algorithm, and an interpreter for split programs.
This implementation extends the language described in \ref{sec:semantics} with 
\rmint{}s,
\rmbool{}s,
\texttt{let} statements,
$n$-ary sums and products,
datatype and function declarations, and
deep pattern matching (including $\next{}$ and $\pure{}$ patterns).
The code snippets in this paper are written in our concrete syntax, using 
these additional features when convenient.
Our expanded syntax allows staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngr{`1`
  datatype list = Empty | Cons of int * list
  fun part (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
val 3`gr{`part3`}` = 3`gr{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

### 2016-01-09 paper/relatedwork.tex
added mixed map and cut cut cut

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is partial
evaluation, which distinguishes between static (compile-time) and
dynamic (runtime) stages. Given a program and values for all static
inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52},
%Kleene did not intend to improve efficiency, however.  Lombardi
%appears to be the first person to use the term in his work on
%incremental computation~\cite{Lombardi67}.
%There has since been an enormous body of work on partial evaluation,
and we refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of partial evaluation work until the early
90's~\cite{JGS93}.

@@

Another class of precomputation techniques is incremental computation,
where a program can efficiently respond to small changes in its inputs
by only recomputing parts of the computation affected by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Thus, incremental computation can be thought of as a staged model,
where changes in inputs come over a period of time in stages.
However, unlike splitting, incremental computation does not require
fixing any of its inputs and, in the general case, allows for all
program inputs to change. Thus, the benefits of incremental
computation depend on what changes to inputs are made. For example,
while it is possible to apply incremental computation to the
quickselect example in \ref{sec:overview}, techniques would unfold the
quickselect function based on the demanded ranks, potentially
incurring linear time cost at each step of the algorithm (as opposed
to the logarithmic result produced by splitting).  Unlike splitting,
incremental computation techniques must also maintain sophisticated
data structures dynamically at run time to track what computations
must be performed.

>>>>

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is partial
evaluation, which distinguishes between static (compile-time) and
dynamic (runtime) stages. Given a program and values for all static
inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  
We refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of partial evaluation work until the early
90's~\cite{JGS93}.

@@

Another class of precomputation techniques is incremental computation,
where a program can efficiently respond to small changes in its inputs
by only recomputing parts of the computation affected by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
However, unlike splitting, incremental computation does not require
fixing any of its inputs and, in the general case, allows for all
program inputs to change. Thus, the benefits of incremental
computation depend on what changes to inputs are made. For example,
while it is possible to apply incremental computation to the
quickselect example in \ref{sec:overview}, techniques would unfold the
quickselect function based on the demanded ranks, potentially
incurring linear time cost at each step of the algorithm (as opposed
to the logarithmic result produced by splitting).  Moreover,
incremental computation techniques must also maintain sophisticated
data structures dynamically at run time to track what computations
must be performed.
